Binary-liquid lattice Boltzmann simulation of microchannels with square cross sections

### 2011-05-30 Text/paper.tex
Merging Schlumberger comments.

\documentclass{article}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\title{Binary-liquid lattice Boltzmann simulation of microchannels with square cross sections}
\author{A.Kuzmin \and M.Januszewski \and D.Eskin \and F.Mostowfi \and J.Derksen}
\begin{document}
\maketitle
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capilary number $Ca$ and Reynolds number $Re$. Recently we investigated
the applicability and parameter choice for the two-dimensional case of the Bretherton problem (flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is a continuation of our previous work.  To validate the applicability
of the LBM to Bretheron phenomena in the moderate range of the capillary number $0.05\leq Ca \leq
6.0$,
we focus on the simulation of three-dimensional capillaries with square cross sections.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to capture a number of phenomena happening in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the dependency
of the bubble radii on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

For example, the deposited film thickness
is proportional to $Ca^{2/3}$ in the range of small capillary numbers for circular channels
\cite{bretherton,heil-bretherton}. 
The problem of predicting flow patterns and associated mass transfers for Bretheron-type flows
is of significant interest for chemical industry as it is widely used in chemical monolith
microreactors \cite{kreutzer-pressure-drop}. Large mass transfers can be achieved because of the
large interfacial area and small diffusion lengths \cite{cerro-bubble-train}. Heat transfer is
also increased in comparison with single phase flow \cite{fukugata-levelset}. While it
is possible to calculate the flow analytically for small capillary numbers and negligible Reynolds
number \cite{bretherton}, the
results cannot be extrapolated to the wider range of capillary and Reynolds numbers used in the
chemical industry.  Thus, the desire for consistent numerical simulations arises.

The two-dimensional geometry (circular tubes, parallel plates) flows are studied extensively in
experimental works of \citet{aussillous-deposition, cerro-bubble-train} and numerical works of
\citet{giavedoni-numerical,heil-bretherton}. All abovementioned works found that the analytical
Bretherton
analysis is valid only for small capillary numbers $Ca\leq 0.003$ and deviates from actual
measurements for larger ones.
This is caused by a complex interplay of gravitational,
interfacial, inertial and viscous forces \cite{gupta-review}. Historically, 
\citet{bretherton} indicated that inertia effects can be 
neglected. \citet{giavedoni-numerical} suggested
that
inertia effects
are negligible for $Ca \leq 0.05$ and have moderate impact for $Ca>0.05$ in the range of Reynolds
number from $0$ to $70$ for two-dimensional flows. Later on, \citet{heil-bretherton}
extended
the results for the flow between plates upto Reynolds number
$300$. This author indicates that while the influence on the established film thickness is
insignificant
($7$ percent deviation from the film thickness measured at $Re=0$), the change in Reynolds number 
influences significantly the pressure distrubution and the flow field near the front bubble tip.
\citet{cerro-bubble-train} show that the mass transfer is strongly affected by flow direction for
small capillary numbers in case of upward and downward flows. Thus, to fully describe the
propagation of the semi-infinite finger in liquid media one needs to take into the account the
viscous, gravitational, surface, and inertia forces \cite{gupta-review}.

In comparison with the two-dimensional Bretherton flows (circular tubes, flow between plates), there
is a vast number of experimental
results available for three-dimensional cases of microchannels with square, triangular and
rectangular cross sections. For instance, \citet{cerro-bubble-train} performed a number of
experiments for a
bubble-train flow in capillaries of
circular and square cross sections for horizontal, upward and downward flows.
\citet{shikazono-square} obtained
experimental
results for the deposition film thickness dependency on the
capillary number for ethanol/air and water/air mixtures and for square, circular and triangular
shaped mirochannels. They also presented the experimental correlation for the prediction of bubble
radii
based on the capillary number $Ca$ and the Weber number $We=Re\,Ca$, where the Reynolds number is
defined as:
\begin{equation}
Re=\frac{U_{\mathrm{bubble}} H_{\mathrm{eff}}}{\nu_{\mathrm{liq}}},
\end{equation}
where $H_{\mathrm{eff}}$ is the channel height and $\nu_{\mathrm{liq}}$ is the kinematic liquid
viscosity.

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} show interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular
or square-shaped capillaries there is a transition from an asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape is attributed to the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
From hereon we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependency
of the diagonal and axial radii on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[ht]
\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The transition capillary number for square channels between non-axisymmetric case and axisymmetric
one is indicated in a number of
works as $\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}. If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

There are also a number of numerical works for the three-dimensional case. For instance, 
\citet{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and calculated bubble shapes for different
slug and channel cross sections.
\citet{heil-threedim} performed three-dimensional simulations for circular-,
square- and rectangular-shaped capillaries. They indicated a transition
capillary number where the streamlines pattern
changes from having a vortex in front of the bubble and not having it for larger capilllary
numbers,  Fig. \ref{fig:streamlines:pattern}. For a square channel the critical
capillary number for this phenomena is $Ca=0.691$.  \citet{heil-threedim} also found an empirical
correlation which
makes it possible to put on one curve the results of radii dependency on the capillary number
simulations of rectangular channels with different aspect ratios.
It is also indicated that for microchannels with a certain aspect ratio $\alpha=\frac{a}{b}\geq
2.04$, the interface does not become
axisymmetric for any capillary number. \citet{wang-non-circular} performed numerical
simulations using the Volume of Fluid (VOF) technique for
capillaries with non-circular (square, triangular) cross sections. They also investigated the relative
velocity of the bubble to the slug liquid velocity  for
a number of capillary numbers.

As it is mentioned by \citet{gupta-review}, the most common techniques to simulate the Bretherton
phenomena are the VOF method \cite{wang-non-circular}, the level-set method
\cite{fukugata-levelset} and the finite element methods \cite{kreutzer-taylor,heil-threedim}. It is
also indicated \cite{gupta-review} that the new techniques which are still in the development stage
are the lattice Boltzmann method and phase field methods \cite{anderson-diffuse,gurtin-binary}.
The finite element method solves the Bretherton flow as a free-surface problem with a sharp
interface but without underlying details for the gas.  The lattice Boltzmann method on the other
hand,
as a continuous interface method, provides more flexibility in simulations involving coalescence
and/or
droplet breakup, and thus arises as a promising alternative for simulation of gas finger propagation.

In our previous work \cite{kuzmin-binary2d} we already covered the Bretherton flow between parallel
plates.
It was shown that the free-energy binary liquid model, which is a phase field method, simulates
with reasonable accuracy the Bretherton/Taylor phenomenon. The goal of this work is to examine 
if similar conclusions can be drawn for three-dimensional flows in microchannels with square cross sections,
simulated using the free-energy binary liquid lattice Boltzmann
method. 
%As it was indicated by \citet{gupta-review} the flow in microchannels is a complex
%interplay of viscous, gravitational, inertial and surface forces. For the sake of comparison, we
%specifically design the numerical benchmark to avoid the influence of gravitational and inertial
%forces.

@@

The binary liquid free-energy LB model due to \citet{swift} we used
simulates two liquids with the assumption of uniform overall
density. As it was indicated by \citet{gupta-review} the flow in microchannels is a complex
interplay of viscous, gravitational, inertial and surface forces, and the classical Bretherton
problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, the parameters in our simulations were carefully chosen to
avoid  inertia
effects, see Section \ref{sec:numerical:benchmark}. Moreover, the numerical results of
\citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effects on the film thickness for a relatively wide range of Reynolds
numbers. Thus, a major governing factor for microchannel flows is not the density ratio, but
the
viscosity ratio, which is why the uniform density binary liquid model is suitable for this kind of
simulations.
% THIS HAS ALREADY BEEN STATED BEFORE
% The goal of this work is to do a feasibility study of the LBM
% binary-liquid model to simulate and correctly predict flow patterns for the Bretherton/Taylor
% problem.

@@

The paper is organized as follows.  First, we explain the simulation benchmark construction
mainly based on our previous work \cite{kuzmin-binary2d}. Then, the binary liquid lattice
Boltzmann model is outlined. The simulation results for the three-dimensional case are presented in
the results section. The results cover the film thickness dependency on the capillary number, the
relative to the slug liquid velocity of the bubble and the vortex profiles. The paper
is concluded with a summary of the main findings.
%The critical capillary numbers for changing velocity pattern depending on the eccentricity
%parameter $\alpha$ are presented in Table \ref{table:recirculation:data}.
%\begin{table}
%\begin{tabular}{c|c}
%$\alpha$& $Ca_{T}$\\
%\hline
%1& 0.691\\
%1.1 & 0.688\\
%1.3 & 0.666\\
%1.5 & 0.631\\
%\end{tabular}
%\caption{The results for the recirculation region.\label{table:recirculation:data}}
%\end{table}

\section{Numerical benchmark approach}
\label{sec:numerical:benchmark}
The main discussion here follows and is based on the two-dimensional benchmark approach
\cite{kuzmin-binary2d}. It was indicated that the benchmark layout should have certain
dimensions to conduct simulations. The classical Bretherton benchmark layout is represented in
Fig.~\ref{fig:classical:benchmark}. It describes the gas finger propagation through the liquid
medium.
The film thickness in this case is measured at the inlet. In the lattice Boltzmann framework such a
formulation has certain challenges \cite{kuzmin-binary2d}. Some of them are attributed to the
dynamic coupling of the inlet and outlet conditions \cite{giavedoni-numerical}. Instead, we
proposed the numerical benchmark indicated in Fig.~\ref{fig:lbm:benchmark}. The dimensions of the
channel are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}} \times 15 H_{\mathrm{eff}}$. The
initial bubble length is taken as $5 H_{\mathrm{eff}}$. \citet{giavedoni-numerical} showed that
the film in the two-dimensional geometries stabilizes at distances of $2.6-4.0$ diameters
from the front tip depending on the Reynolds number. \citet{heil-threedim} measured the bubble radii
at the distance of $5.5 H_{\mathrm{eff}}$ and indicated it to be sufficient for $Ca\leq 10$. In
this work the following relation holds for all conducted simulations: $0.05 \geq Ca \leq 6$.
Thus, the
film thickness is chosen to be measured in the
middle of the bubble, which is located at least at a distance of $2.5-3$ channel heights from the 
bubble tip.%at the distances one channel height from the rear tip upto one channel

For simplicity, periodic boundary conditions are applied so that effectively a bubble train is simulated
instead of a single bubble.  The mutual influence of bubbles on each other is minimized by the use
of a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which require a
simulataneouse imposement of pressure boundary conditions and the body force. The pressure boundary
conditions for the binary liquid model is a challenge. Thus, it is assumed the flow is in the
horizontal direction only and the upward/downward flows with pressure difference imposed are not
simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{Figures/benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semiinfinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{Figures/benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

@@

\citet{heil-threedim} specify that the diagonal bubble radius is different from the
axial bubble radius for $Ca\leq \widehat{Ca}\approx 0.05$. The radii are the same for larger values
of
$Ca$ with a value $R_{diag}=R_{axis}=0.49 H_{\mathrm{eff}}$. Therefore, taking the minimal
requirement for the film thickness to be
resolved, i.e. $6$ lattice nodes, one would obtain a grid size of $600\times 600 \times 9000 $ to
properly resolve the binary liquid flow. This size implies that relatively
large
computational resources are required to perform the simulations.

@@

% TODO: This paragraph should be revisited, somehow it reads a little weirdly..
It should be noted that the binary liquid model we used has a uniform density. A part of discussion is
addressed in the introduction that to fully address the Bretherton problem one needs to take into
the account gravitational, viscous, inertial, and surface tension forces. Among all the works there
is no full 
agreement between results even in the limit of Reynolds number $0$.  Moreover, as it was indicated
by \citet{kreutzer-taylor} and by \citet{cerro-bubble-train} the pressure distribution, streamlines
and bubble shape are strongly affected by the length of the liquid slug and the length of the
bubble. In current simulations we minimize the effects of bubbles train and inertia to have a least
impact, as the largest Reynolds number is not bigger than $10$ and the lengths of a bubble and a
channel are chosen to be as to minimize mutual effects of bubbles motion on each other and for the
film thickness to stabilize. However, more thorough studies are needed in order to understand
precisely the abovementioned effects.

@@

\subsection{Steady-state approach}
\label{sec:steady:state}
We performed different simulations to understand the number of time steps required for the
simulation to
settle down to a steady state. The grid to be simulated is $52 \times 52 \times 750$ which simulates
a quarter of the channel with the initial film thickness of $6.5$ lattice Boltzmann units together
with the body force $\frac{\mathrm{d}P}{\mathrm{d}x}=1.6
\cdot 10^{-6}$. The simulation was performed for $140000-250000$ iterations with the step of
$10000$ iterations. The results are summarised in Table \ref{table:steady:state}. While the
capillary number variation is around $5$ percent, the radii variation is $0.1$ percent. 
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}
\hline
$N_{iter}$&$Ca$&$R_{axis}$&$R_{diag}$\\
\hline
$140000$&$0.1879$&$0.8657$&$0.8708$\\
$150000$&$0.1879$&$0.8652$&$0.8704$\\
$160000$&$0.1878$&$0.8650$&$0.8702$\\
$170000$&$0.1880$&$0.8648$&$0.8701$\\
$180000$&$0.1882$&$0.8647$&$0.8700$\\
$190000$&$0.1885$&$0.8646$&$0.8699$\\
$200000$&$0.1893$&$0.8645$&$0.8698$\\
$210000$&$0.1963$&$0.8645$&$0.8697$\\
$220000$&$0.1925$&$0.8644$&$0.8697$\\
$230000$&$0.1936$&$0.8644$&$0.8697$\\
$240000$&$0.1961$&$0.8645$&$0.8697$\\
$250000$&$0.1932$&$0.8645$&$0.8698$\\
\hline
\end{tabularx}
\caption{Results for the steady-state case. One
can see that $200000$ steps are enough to shape
the bubble and approach the steady state. The small noise in the capillary number is connected to
the identification of the interface. One needs to address the interpolation and the spurious
currents existing in the system. All other parameters are relaxed to the
steady-state.\label{table:steady:state}}
\end{table}

>>>>

\documentclass{article}
\usepackage[numbers,sort&compress]{natbib}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}
\usepackage{color}
\usepackage{tabularx}
\usepackage{overpic}
\title{Binary-liquid lattice Boltzmann simulation of microchannels with square cross sections}
\author{A.Kuzmin \and M.Januszewski \and D.Eskin \and F.Mostowfi \and J.Derksen}
\begin{document}
\maketitle
\begin{abstract}
The classical Bretherton problem describes the propagation of gas fingers through liquid media in a
narrow channel with
thin liquid films between bubbles and channel walls. The bubble shape and flow patterns are
complicated functions of the capilary number $Ca$ and Reynolds number $Re$. Recently, we
investigated
an applicability and parameter selection for the two-dimensional case of the Bretherton problem
(flow
between parallel plates) using the free-energy binary liquid lattice Boltzmann method (LBM)
\cite{kuzmin-binary2d}. This paper is a continuation of our previous work.  To validate an
applicability
of the LBM to Bretheron phenomena in the moderate range of the capillary number $0.05\leq Ca \leq
6.0$,
we focus on the simulation of three-dimensional capillaries with square cross sections.
The flow is driven by a body force, and
periodic boundary conditions are applied in the streamwise direction. The results show that the
binary liquid model is able to capture a number of phenomena occurin in three-dimensional
capillaries, such as the existence of a vortex in front of the bubble and the dependency
of the bubble radii on the capillary number.  We conclude that lattice Boltzmann free energy binary
liquid model can be used to simulate the Bretherton problem with good accuracy. 
\end{abstract}

@@

For example, the film thickness
is proportional to $Ca^{2/3}$ in the range of small capillary numbers for circular channels
\cite{bretherton,heil-bretherton}. 
The problem of predicting flow patterns and associated mass transfers for Bretheron-type flows
is of significant interest for chemical industry as it is widely used in chemical monolith
microreactors \cite{kreutzer-pressure-drop}. Intensive mass transfer can be achieved because of the
large interfacial area and small diffusion lengths \cite{cerro-bubble-train}. Heat transfer is
also increased in comparison with single phase flow \cite{fukugata-levelset}. While it
is possible to calculate the flow analytically for small capillary numbers and nearly zero Reynolds
number \cite{bretherton}, the
results cannot be extrapolated to the wide range of capillary and Reynolds numbers used in the
chemical industry.  Thus, the need of consistent numerical simulations arises.

The two-dimensional geometry (circular tubes, parallel plates) flows are studied extensively in
experimental works of \citet{aussillous-deposition, cerro-bubble-train} and numerical works of
\citet{giavedoni-numerical,heil-bretherton}. All abovementioned investigations found that the 
Bretherton
analysis is valid only for small capillary numbers $Ca\leq 0.003$ and deviates from actual
measurements for larger ones.
This is caused by a complex interplay of gravitational,
interfacial, inertial and viscous forces \cite{gupta-review}. Historically, 
\citet{bretherton} neglected inertia effects. \citet{giavedoni-numerical} suggested
that
inertia effects
are negligible for $Ca \leq 0.05$ and have moderate impact for $Ca>0.05$ in the range of Reynolds
number from $0$ to $70$ for two-dimensional flows. Later on, \citet{heil-bretherton}
extended
the results for the flow between plates up to $Re=300$. This author noted that while the
influence of inertia on the established film thickness is
insignificant
($7$ percent deviation from the film thickness measured at $Re=0$), the change in Reynolds number
significantly influences  the pressure distrubution and the flow field near the front bubble cap.
\citet{cerro-bubble-train} showed that the mass transfer is strongly affected by flow direction for
small capillary numbers in case of upward and downward flows. Thus, to fully describe the
propagation of the semi-infinite finger in liquid media one needs to take into account the
viscous, gravitational, surface, and inertia forces \cite{gupta-review}.

In comparison with the two-dimensional Bretherton flows (circular tubes, flow between plates), there
is a vast number of experimental
results available for three-dimensional flows in microchannels with square, triangular and
rectangular cross sections. For instance, \citet{cerro-bubble-train} performed a number of
experiments for a
bubble-train flow in capillaries of
circular and square cross sections for horizontal, upward and downward flows.
\citet{shikazono-square} obtained
experimental
results for the film thickness dependence on the
capillary number for ethanol/air and water/air mixtures and for square, circular and triangular
shaped mirochannels. They also presented the experimental correlation for  bubble
radii
based on the capillary number $Ca$ and the Weber number $We=Re\,Ca$, where the Reynolds number is
defined as:
\begin{equation}
Re=\frac{U_{\mathrm{bubble}} H_{\mathrm{eff}}}{\nu_{\mathrm{liq}}},
\end{equation}
where $H_{\mathrm{eff}}$ is the channel height and $\nu_{\mathrm{liq}}$ is the kinematic liquid
viscosity.

The experimental works
\cite{shikazono-square,cerro-bubble-train} supported by numerical simulations \cite{heil-threedim,
wang-non-circular} showed interesting phenomena in three-dimensional geometries. It was found
\cite{heil-threedim,wong-films} that for rectangular
or square-shaped capillaries there is a transition from an asymmetric to an axisymmetric shape for a
certain capillary number.
The non-axisymmetric shape is attributed to the wall normal radius (in Fig.
\ref{fig:lbm:benchmark} the wall normal radius is measured along $x$ or $y$ axes) being 
different from the
diagonal radius and the bubble having a non-circular shape in the channel crossection. In this case,
the bubble shape mimics the shape of a rectangle and looks like a rectangle with rounded corners.
Further we adapt the convention ``axial`` for the radius measured along the $x$ or $y$ axes,
see Fig. \ref{fig:lbm:benchmark} \cite{heil-threedim}. The dependence
of the diagonal and axial radii on the capillary number is shown in Fig. 
\ref{fig:heil:three:dim}.
\begin{figure}[ht]
\begin{overpic}[width=\textwidth]{Figures/capillary_width_heil.eps}
\put(50,45){\includegraphics[width=0.4\textwidth]{Figures/inset.eps}}
\end{overpic}
%\includegraphics[width=\textwidth]{Figures/capillary_width_heil.eps}
\caption{\citet{heil-threedim} results for the variation of the bubble radii for a range of
capillary numbers for a square channel. One can see the asymmetry between diagonal and axial
diameters for the capillary
numbers $Ca\leq\widehat{Ca}=0.04$. Courtesy of \citet{heil-threedim}. \label{fig:heil:three:dim}}
\end{figure}
The capillary number for square channels indicating transition between non-axisymmetric case to
axisymmetric
one is reported in a number of
works ($\widehat{Ca}=0.04$ \cite{cerro-bubble-train},
$\widehat{Ca}=0.1$
\cite{cerro-space,wang-non-circular}, $\widehat{Ca}=0.033$ \cite{heil-threedim}). If the capillary
number is larger
than
the critical capillary number, i.e. $Ca>\widehat{Ca}$, then the bubble becomes axisymmetric with the
radius of the bubble dependent on the capillary number. An example for the bubble radii
dependence on the capillary number is presented in Fig. \ref{fig:heil:three:dim}.

There are also a number of numerical works on three-dimensional flows. For instance, 
\citet{wong-films,wong-pressure} studied
three-dimensional bubbles in
polygonal capillaries and calculated bubble shapes for different
slug and channel cross sections.
\citet{heil-threedim} performed three-dimensional simulations for circular-,
square- and rectangular-shaped capillaries. They indicated a transition of flow regime in a liquid
slug. Vortexes observed at lower capillary numbers disappear at the capillary number above a
certain threshold,  Fig. \ref{fig:streamlines:pattern}. For a square channel the critical
capillary number is $Ca=0.691$.  \citet{heil-threedim} also found an empirical
correlation which
makes it possible to put on the same curve the results of radii dependency on the capillary number
simulations of rectangular channels with different aspect ratios.
It was also found that for microchannels with a certain aspect ratio $\alpha=\frac{a}{b}\geq
2.04$, the interface does not become
axisymmetric for any capillary number. \citet{wang-non-circular} performed numerical
simulations using the Volume of Fluid (VOF) technique for
capillaries with non-circular (square, triangular) cross sections. They also investigated the
relative slug to bubble velocity for
a range of capillary numbers.

As it was mentioned by \citet{gupta-review}, the most common techniques to simulate the Bretherton
phenomena are the VOF method \cite{wang-non-circular}, the level-set method
\cite{fukugata-levelset} and the finite element methods \cite{kreutzer-taylor,heil-threedim}. It was
also indicated \cite{gupta-review} that the new techniques, which are still in the development
stage,
are the lattice Boltzmann method and phase field methods \cite{anderson-diffuse,gurtin-binary}.
The finite element method solves the Bretherton flow as a free-surface problem with a sharp
interface but without underlying details for the gas.  The lattice Boltzmann method is a continuous
interface method, and therefore provides more flexibility in simulations involving coalescence
and/or
droplet breakup, and thus arises as a promising alternative for simulation of gas finger propagation.

In our previous work \cite{kuzmin-binary2d}, we already analyzed the Bretherton flow between
parallel
plates.
It was shown that the free-energy binary liquid model, which is a phase field method, simulates the
Bretherton/Taylor phenomenon with reasonable accuracy. The goal of this work is to examine 
if similar conclusions can be drawn for three-dimensional flows in microchannels with square cross
sections,
simulated using the free-energy binary liquid lattice Boltzmann
method. 
%As it was indicated by \citet{gupta-review} the flow in microchannels is a complex
%interplay of viscous, gravitational, inertial and surface forces. For the sake of comparison, we
%specifically design the numerical benchmark to avoid the influence of gravitational and inertial
%forces.

@@

The binary liquid free-energy LB model due to \citet{swift} we used
simulates two liquids with the assumption of uniform overall
density. As it was indicated by \citet{gupta-review} a flow in microchannels is a complex
interplay of viscous, gravitational, inertial and surface forces, and the classical Bretherton
problem is stated for gas and liquid, which are of
significantly
different densities and viscosities, the parameters in our simulations were carefully chosen to
avoid  inertia
effects, see Section \ref{sec:numerical:benchmark}. Moreover, the numerical results of
\citet{giavedoni-numerical} and \citet{heil-bretherton} show
negligible Reynolds number effects on the film thickness for a relatively wide range of Reynolds
numbers. Thus, a major governing factor for microchannel flows is not the density ratio, but
the
viscosity ratio. Therefore the uniform density binary liquid model is suitable for this kind of
simulations.
% THIS HAS ALREADY BEEN STATED BEFORE
% The goal of this work is to do a feasibility study of the LBM
% binary-liquid model to simulate and correctly predict flow patterns for the Bretherton/Taylor
% problem.

@@

The paper is organized as follows.  First, we explain the simulation benchmark construction
mainly based on our previous work \cite{kuzmin-binary2d}. Then, the binary liquid lattice
Boltzmann model is outlined. The simulation results for the three-dimensional case are presented in
the results section. That covers the film thickness dependency on the capillary number, the bubble
to slug relative velocity and vortex profiles. The paper
is concluded with a summary of the main findings.
%The critical capillary numbers for changing velocity pattern depending on the eccentricity
%parameter $\alpha$ are presented in Table \ref{table:recirculation:data}.
%\begin{table}
%\begin{tabular}{c|c}
%$\alpha$& $Ca_{T}$\\
%\hline
%1& 0.691\\
%1.1 & 0.688\\
%1.3 & 0.666\\
%1.5 & 0.631\\
%\end{tabular}
%\caption{The results for the recirculation region.\label{table:recirculation:data}}
%\end{table}

\section{Numerical benchmark approach}
\label{sec:numerical:benchmark}
The main discussion here is based on the two-dimensional benchmark approach
\cite{kuzmin-binary2d}. It was indicated that the benchmark layout should have certain
dimensions to conduct simulations. The classical Bretherton benchmark layout is represented in
Fig.~\ref{fig:classical:benchmark}. It describes the gas finger propagation through the liquid
medium.
The film thickness in this case is measured at the inlet. In the lattice Boltzmann framework, such a
formulation has certain challenges \cite{kuzmin-binary2d}. Some of them are attributed to the
dynamic coupling of the inlet and outlet conditions \cite{giavedoni-numerical}. Instead, we
proposed the numerical benchmark indicated in Fig.~\ref{fig:lbm:benchmark}. The dimensions of the
channel are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}} \times 15 H_{\mathrm{eff}}$. The
initial bubble length is taken as $5 H_{\mathrm{eff}}$. \citet{giavedoni-numerical} showed that
the film in the two-dimensional geometry stabilizes at distances of $2.6-4.0$ diameters
from the front tip depending on the Reynolds number. \citet{heil-threedim} measured the bubble radii
at the distance of $5.5 H_{\mathrm{eff}}$ and indicated it to be sufficient for $Ca\leq 10$. In
this work, the following relation holds for all conducted simulations: $0.05 \geq Ca \leq 6$.
Thus, the
film thickness is chosen to be measured in the
middle of the bubble, which is located at least at a distance of $2.5-3$ channel heights from the 
bubble tip.%at the distances one channel height from the rear tip upto one channel

For simplicity, periodic boundary conditions are applied. Therefore  a bubble
train is simulated
instead of a single bubble.  The mutual influence of neighbouring bubbles is minimized
by using a long
channel ($15 H_{\mathrm{eff}}$). Periodic boundary conditions imply as well that the fluid is driven
by a
body force in the framework of the LBM and not with a pressure difference. Therefore, we can not
address simulations with upward and downward flows \cite{cerro-bubble-train} which require a
simulataneous imposement of pressure boundary conditions and the body force. The pressure boundary
conditions for the binary liquid model is a challenge. Thus, we consider a flow in the
horizontal direction only and the upward/downward flows with pressure difference imposed are not
simulated. 
\begin{figure}[htb!]
\includegraphics*[bb=153 610 410 717,width=\textwidth]{Figures/benchmark_classical.eps} 
\caption{The classical benchmark layout describes a semiinfinite gas bubble
propagation through the liquid media. The inlet pressure is specified $P_{in}$, the outlet is the
free outflow. The bubble propagates with constant velocity $U_{\mathrm{bubble}}$. 
\label{fig:classical:benchmark}}
\end{figure}
\begin{figure}[htb!]
\includegraphics*[bb=152 470 415 713,width=\textwidth]{Figures/benchmark_lbm.eps}
\caption{The lattice Boltzmann benchmark layout as used in present calculations. The dimensions of
the domain are chosen as $H_{\mathrm{eff}}\times H_{\mathrm{eff}}\times 15 H_{\mathrm{eff}}$. The
length of the bubble was chosen as $5 H_{\mathrm{eff}}$ for the film thickness to stabilize.
Periodic boundary conditions are applied in the streamwise direction. The flow is driven by a body
force. \label{fig:lbm:benchmark}}
\end{figure}

@@

\citet{heil-threedim} specify that the diagonal bubble radius is different from the
axial bubble radius for $Ca\leq \widehat{Ca}\approx 0.03$. The radii are the same for larger values
of
$Ca$ with a value $R_{diag}=R_{axis}=0.49 H_{\mathrm{eff}}$. Therefore, taking the minimal
requirement for the film thickness to be
resolved, i.e. $6$ lattice nodes, one would obtain a grid size of $600\times 600 \times 9000 $ to
properly resolve the binary liquid flow. This size implies that relatively
large
computational resources are required to perform the simulations.

@@

% TODO: This paragraph should be revisited, somehow it reads a little weirdly..
It should be noted that the binary liquid model, we used, has a uniform density. A part of
discussion is
addressed in the introduction:  to describe the Bretherton problem one needs to take into
account gravitational, viscous, inertial, and surface tension forces. Among all the works, there
is no full 
agreement between results even in the limit of Reynolds number $0$.  Moreover, as it was indicated
by \citet{kreutzer-taylor} and by \citet{cerro-bubble-train} the pressure distribution, streamlines
and bubble shape are strongly affected by the length of the liquid slug and the length of the
bubble. In current simulations we minimize the effects of bubbles train and inertia to have a least
impact, as the largest Reynolds number is not bigger than $10$ and the lengths of a bubble and a
channel are chosen to be as to minimize mutual effects of bubbles motion on each other and for the
film thickness to stabilize. However, more thorough studies are needed in order to understand
precisely the abovementioned effects.

@@

\subsection{Steady-state approach}
\label{sec:steady:state}
We performed different simulations to understand the number of time steps required for the
simulation to
settle down to a steady state. The grid  was $52 \times 52 \times 750$ which represents
a quarter of the channel with the initial film thickness of $6.5$ lattice Boltzmann units together
with the body force $\frac{\mathrm{d}P}{\mathrm{d}x}=1.6
\cdot 10^{-6}$. The simulation was performed $250000$ iterations with the step of
$10000$ iterations. The results for time iterations $140000-250000$ are summarised in Table
\ref{table:steady:state}. While the
capillary number variation is around $5$ percent, the radii variation is $0.1$ percent. 
\begin{table}
\begin{tabularx}{\textwidth}{|X|X|X|X|}
\hline
$N_{iter}$&$Ca$&$R_{axis}$&$R_{diag}$\\
\hline
$140000$&$0.1879$&$0.8657$&$0.8708$\\
$150000$&$0.1879$&$0.8652$&$0.8704$\\
$160000$&$0.1878$&$0.8650$&$0.8702$\\
$170000$&$0.1880$&$0.8648$&$0.8701$\\
$180000$&$0.1882$&$0.8647$&$0.8700$\\
$190000$&$0.1885$&$0.8646$&$0.8699$\\
$200000$&$0.1893$&$0.8645$&$0.8698$\\
$210000$&$0.1963$&$0.8645$&$0.8697$\\
$220000$&$0.1925$&$0.8644$&$0.8697$\\
$230000$&$0.1936$&$0.8644$&$0.8697$\\
$240000$&$0.1961$&$0.8645$&$0.8697$\\
$250000$&$0.1932$&$0.8645$&$0.8698$\\
\hline
\end{tabularx}
\caption{Results for the steady-state case. One
can see that $200000$ steps are enough to shape
the bubble and approach the steady state. The small noise in the capillary number is connected to
the identification of the interface. One needs to address the interpolation and the spurious
currents existing in the system. All other parameters are relaxed to the
steady-state.\label{table:steady:state}}
\end{table}
